---
title: "Current and Future Deforestation in Amazon Forest"
author: "Jiawen Tang, Xiangrong (Mark) Sun"
maketitle: true
output: github_document
---

Amazon Forest is the world's largest tropical rain forest located in the south america. Known as the lung of earth, it is the largest single carbon storage system and the largest terrestrial oxygen producer. However, this huge rain forest is facing large scale deforestation and degradation due to human activities. In this project, we will focus on Brazil's part of deforestation, finding the historical data to understand deforestation status and projecting the future.

We will be using several data sources, and replicate data from research paper to show the historical and current status of deforestation. We will also explore future forest conditions under different scenarios. The main paper we aim to replicate is <https://www.science.org/doi/10.1126/science.abp8622>, we will start with its current estimate and end with this as well. Apart from it, we will use other data sources such as fire to establish connections with deforestation.

Starting with the study, we will use several functions to create graphs, maps, reform and visualize the data. We get in touch with multiple use of dplyr, ggplot, sf, tmap, tidyverse, and we bring in new packages and functions such as rnaturalearth. Before we start, we will install and read all the packages we need for the study:

```{r setup, messages = FALSE, cache = FALSE, warning = FALSE}
#install tmap to upgrade it each time when reopen the file
suppressMessages({
  install.packages(c('tmap','rnaturalearth','rnaturalearthhires'))
  library(fs)
  library(terra)
  library(tidyverse)
  library(tmap)
  library(rnaturalearth)
  library(sp)
  library(sf)
  library(reshape2)
  library(dplyr)
  library(ggplot2)
})
```

## Historical Deforestation

We will begin with reproducing Figure 2 of our main paper, the 2001-2018 spatial distribution of four major drivers of Amazonian forest degradation, excluding deforestation and savanna areas. We start with reading the tif file to raster and using the r-data 'World' as well as "tmap" to create maps. The four major drivers are fire burning, extreme drought occurrence, area within a forest edge, and timber extraction.

```{r}
#read in the tif and World
fire <- rast("fire.tif")
drought <- rast("drought.tif")
edge <- rast("edge.tif")
logging <- rast("logging.tif")
data("World")

#Create maps by changing making sure all data are visualizable. 
fire[fire == 0] <- NA
drought[drought == 0] <- NA
edge[edge == 0] <- NA
logging[logging == 0] <- NA
tm_shape(World,bbox = stars::st_as_stars(fire))+tm_polygons() +
  tm_shape(fire)+tm_raster()
tm_shape(World,bbox = stars::st_as_stars(drought))+tm_polygons() +
  tm_shape(drought)+tm_raster()
tm_shape(World,bbox = stars::st_as_stars(edge))+tm_polygons() +
  tm_shape(edge)+tm_raster()
tm_shape(World,bbox = stars::st_as_stars(logging))+tm_polygons() +
  tm_shape(logging)+tm_raster()
```

## Brazil Deforestation in Amazon Forest

To get a direst, overall sense of the geographical distribution of deforestation, we will use state data from INPE (National Institute for Space Research in Brazil National Penitentiary Institute of Peru) of Brazil to visualize forest loss of different states. First we read theis Brazilian deforestation data file, which contains deforested area within 9 states, as well as a total forest lost, from 2004 to 2019.

```{r}
deforest <- read.csv("def_area_2004_2019.csv")
colnames(deforest)= c("Year","Acre","Amazonas","Amapa","Maranhao","Mato Grosso","Para","Rondonia","Roraima","Tocantins","Total")
names(deforest)
```

We want to plot out each states' lost of forest throughout these years, so we can decide how to visualize forest lost on map. For now, we present the data with a line plot:

```{r}
deforested <- melt(deforest,id="Year")
deforest_show <- ggplot(deforested,aes(x=Year,y=value,colour=variable,group=variable)) + geom_line()
deforest_show + labs(title = "Deforestation in different states of Amazon") + labs(x = "Year") + labs(y = "Deforested Area in"~km^2)
```

At this point, we want to see the percentage lost and the total lost of each state over these years, so we can create maps for them. We will calculate the sum of deforested area for each state, and read in a new data from world bank. We also create a list of area code for future reference.

```{r}
Statename = c("Acre","Amazonas","Amapá","Maranhão","Mato Grosso","Pará","Rondônia","Roraima", "Tocantins")
Sumofdef <- c(sum(deforest$Acre),sum(deforest$Amazonas),sum(deforest$Amapa),sum(deforest$Maranhao),sum(deforest$`Mato Grosso`),sum(deforest$Para),sum(deforest$Rondonia),sum(deforest$Roraima),sum(deforest$Tocantins))
areacode <- c("BR-AC","BR-AM","BR-AP","BR-MA","BR-MT","BR-PA","BR-RO","BR-RR","BR-TO") 
forest <- read.csv("2020Forest.csv")
forest
```

Now, we will create a table that contains all information together. Since when we create them they are in correct order, we will just frame them together, merge in the deforestation data and mutate the lost percentage.

```{r, messages = FALSE, cache = FALSE, warning = FALSE}
total_deforest <- data.frame(state = Statename, iso_3166_2 = areacode,
                   Area = Sumofdef)
total_deforest <- merge(total_deforest, forest, by = "iso_3166_2", all.x = TRUE)
total_deforest  <- total_deforest %>%
  mutate(lost_percentage = (Area / CurrentArea) * 100)
total_deforest
```

We will visualize the table by graph, one bar-chart for the deforestation area, and another dot-chart for the deforestation percentage:

```{r}
ggplot(total_deforest, aes(x = state, y = Area)) +
  geom_bar(stat = "identity", fill = "brown") +
  labs(title = "Total deforested areas in Brazil", x = "States", y = "Area (km^2)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplot(total_deforest, aes(x = iso_3166_2, y = lost_percentage, group = 1)) +
  geom_point(color = "red", size = 3) +
  geom_line() +
  labs(title = "Forest Lost Percentage Over States", x = "State Code", y = "Lost Percentage")
```

From the above two charts, we can clearly see that the area and ratio of deforestation are different. However, since these names cannot be expressed concretively for most non-locals, we will plot them on the map of Brazil. We will use the new function, rnaturalearth to find out the area we want. This function is useful but need to re-install every time refreshing the code. Please note this could causing issue when run without internet and without response to the console. Since the function will have a very long result (more than 2000 lines), we will just present the states code and the area to check if its correct.

```{r}
#response '1' and '1' for both install question if needed.
states <- ne_states(country = "Brazil")
states <- st_as_sf(states)
final <- inner_join(states,total_deforest,by = "iso_3166_2" )

```

Now we can start drawing the map. We have the base map and the data, and we plot it using tmap. From the graph we cansee that Para is the main forest-loss state, but due to its large area, the ratio is not very high. In contrast, the neighboring marginal provinces experienced forest losses of more than 10 per cent because they had less forest themselves. Because of the data issue, we only can use the forest area of 2020, which will increase the percentage since 2000 will have a larger forest area for most of the states.

```{r}
tm_shape(final) + tm_polygons("Area") + tm_symbols + tm_layout(legend.outside=TRUE) + tm_text("state", size = 0.5) 
tm_shape(final) + tm_polygons("lost_percentage") + tm_symbols + tm_layout(legend.outside=TRUE) + tm_text("state", size = 0.5)
```

## Fire frequency in Brazil

Fire is also a major driver of forest lost in Amazon. In recent years, destructive wildfire has increased both frequency and intensity due to climate change and other human disturbances, resulting in massive forest loss. In this section, we use a fire data from Kaggle to visualize the frequency of fire occurrences in different states of Brazil over the years, and conclude geo-spatial trends.

First, we read this fire data, which we made some of our own edition.

```{r}
firedata <- read.csv("amazon.csv")
names(firedata)
```

We first group the fire data by "year", and calculate the mean number of fires per year in Brazil from 1998 to 2017.After that, we plot the trend using line plot. From the plot, it seems that the mean number of fire events in Brazil has been rising through time, despite having some fluctuations.

```{r}
MeanBrazil <- firedata %>%
  group_by(year) %>%
  summarize(mean_value = mean(number))
ggplot(MeanBrazil, aes(x = year, y = mean_value)) +
  geom_line() +
  labs(title = "Mean fire occurences in Brazil", x = "year", y = "mean fire occurences")     
```

Now we are interested in how does fire frequency varies between different states in Brazil. We group by "state" column and show the result as bar-chart. There are big geo-spatial variations of fire frequency in Brazil, which we will visualize as map in later steps.

```{r}
MeanState <- firedata %>%
  group_by(state) %>%
  summarize(meanfire = mean(number))
MeanState$state <- str_replace(MeanState$state, "\t", "")
ggplot(MeanState, aes(x = state, y = meanfire)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mean Fire by State", x = "State", y = "Mean Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We also want to see the monthly variations in fire occurrences of all states and time, so we group the data set by "month" column and summarize the mean. Since the original data set has month names in Portuguese, and order Alphabetically by default instead of from Jan to Dec, we made some adjustments. From the final bar-chart, it's obvious that the second half of the year (from July to December) has significantly more fire events on average than the first half.

```{r}
MeanMonth <- firedata %>%
  group_by(month) %>%
  summarize(mean_value = mean(number))
MeanMonth$month = c("April","August","December","Feburary","January","July","June","May","March","November","October","September")
MeanMonth

MMdf_new <- data.frame(
  Month= month.name,
  Mean = c(88.25849,57.12602,56.88408,52.20143,64.31734,103.72347,170.97428, 163.05636,108.47834,164.22515,158.34825,112.15493)
)
MMdf_new$Month <-factor(MMdf_new$Month, levels = c("January", "February", "March", "April",
                                        "May", "June", "July", "August",
                                        "September", "October", "November", "December"))
MMdf_new

ggplot(MMdf_new, aes(x = Month, y = Mean)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mean Values by Month", x = "Month", y = "Mean Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Here, we join our fire data with state-code information, and visualize the geo-spatial distribution of fire events within Brazil using tmap. The most frequent fire events happen in BR-MT (Mato Grosso) and BR-SP (São Paulo).It's highly likely that all states will experience more fire events in the future, showing a redder map by then.

```{r}
mapfire <- firedata |> group_by(iso_3166_2) |>
  summarize(meanfire = mean(number))
mapfire$iso_3166_2 <- str_replace(mapfire$iso_3166_2, "^\\t", "")
firemap <- inner_join(states,mapfire,by = "iso_3166_2" )
tm_shape(firemap) + tm_polygons("meanfire") + tm_symbols + tm_layout(legend.outside=TRUE) + tm_text("iso_3166_2", size = 0.5) +tmap_options(check.and.fix = TRUE)
```

## Drivers of forest-loss in Brazil amazon forest

On a broader scale, we are concerned about the effects of multiple drivers of forest loss in Brazil amazon. We are able to do that using data collected by Tyukavina et al. (2017). We read and pivot the data file, and visualize it using geom_area(). Overall, there is a decreasing trend of forest loss over time, and pasture has always been the predominant cause of forest loss.

```{r}
driver <- read.csv("drivers-forest-loss-brazil-amazon.csv")
driver <- driver %>% select(-Entity)
driver <- driver %>% select(-Code)
driver_year <- driver |> pivot_longer(-Year)
driver_year
ggplot(driver_year,aes(x=Year,y=value,fill=name)) + geom_area()+ labs(x = "Year", y = "forest loss in hectare")
```

## Future Deforestation

Let's go back to the future mentioned in the original paper. We show that fire frequency is expected to increase in response to human activity and climate change, along with continuing logging, under the "Business-as-usual" scenario.These activities will exacerbate the ecological crisis in Amazon, worsening the ecosystem health and integrity.

```{r}
baulogg <-rast("bau_logg_final.tif") 
baufire <-rast("bau_fire_final.tif")
baulogg[baulogg < 0] <- NA
baufire[baufire < 0] <- NA
tm_shape(World, bbox = stars::st_as_stars(baulogg), raster.downsample = list(width = 1140, height = 877)) +
    tm_polygons() +
    tm_shape(baulogg) + tm_raster()
tm_shape(World, bbox = stars::st_as_stars(baufire), raster.downsample = list(width = 1140, height = 877)) +
    tm_polygons() +
    tm_shape(baufire) + tm_raster()
```

## Conclusion

In this project, we were able to analyzed data and create figures to interpret the ecological crisis on Amazon forest in Brazil. We recreated maps showing the spatial distribution of major drivers of Amazonian forest degradation. We show the geo-spatial and temporal trends in deforestation of Amazon forest in Brazil. We then delved into a major driver of forest-loss: fire, and summarized its trend for various places and times. Lastly, we recreated maps showing future projection of logging and fire events in Brazil, which are expected to accelerate forest loss.

With all the data and maps, we clearly see that forest-loss has been increasing in scale through years, along with some major drivers like fire. Understanding their pattern is crucial for humans to make reasonable conservation decisions to reverse the current dire trend.

## Reference

Tyukavina et al. (2017). Types and rates of forest disturbance in Brazilian Legal Amazon, 2000--2013. Science

<https://www.kaggle.com/datasets/mbogernetto/brazilian-amazon-rainforest-degradation/data>

<https://www.kaggle.com/code/instalok/brazil-forest-fire/input>

<https://data.worldbank.org/indicator/AG.LND.FRST.ZS?locations=BR>
